Plasticity and adaptation to high light intensity amplify the advantage of amphistomous leaves

### 2025-07-27 ms/ms.qmd
draft sent for feedback

The presence of stomata on both leaf surfaces (amphistomy) can increase photosynthesis in C$_3$ plants by reducing the path length for CO$_2$ diffusion between substomatal cavities and chloroplasts. Amphistomatous leaves are common among herbaceous plants growing in sunny habitats, including many crop species. This distribution of amphistomatous leaves in nature may result from an increased photosynthetic benefit of amphistomy under high light intensity, either because of acclimatory, plastic, or constitutive variation in CO$_2$ supply or demand. We used a recently developed method to quantify amphistomy advantage, the photosynthetic rate of an amphistomatous leaf relative to an otherwise identical hypostomatous leaf, in `r n_acc` diverse populations representing `r n_sp` species of wild tomatoes (*Solanum* sect. *Lycopersicon* and sect. *Lycopersicoides*). Plants grown under high light intensity benefit more in terms of CO$_2$ assimilated for a given stomatal conductance than plants grown under low light, regardless of measurement light intensity. Furthermore, populations native to open habitats tend to benefit more from amphistomy compared to those from more closed habitats. Developmental plasticity and adaptation of leaves to high light intensity likely play a combined role in explaining the adaptive significance and distribution of amphistomy in nature.

@@

The magnitude of \aax{} we document across populations is also noteworthy, as it implies that hypostomatous leaves predominant in mesic to wet forests globally are giving up 'free' carbon that would require little to no additional water loss. Back-of-the-envelope calculations based on our data suggest that leaves could use X-X% less water for the same carbon gain if they were amphistomatous (see Supporting Information for details of calculations). As this is the first quantitative comparison of \aax{} across species, estimation from a broader range of species and environments will be required to fully understand the ecological causes and consequences of stomatal distributions on leaf surfaces.

@@

To test whether LMA mediated the relationship between growth light intensity and \aax{}, we repeated a similar model comparison procedure but included $\log \mathrm{LMA}$ as an additional fixed effect in the model, as well its interaction with measurement and growth light intensity. The model included a direct effect of growth light intensity on $\log \mathrm{LMA}$ and random phylogeneticlly structured effects of population on $\log \mathrm{LMA}$ and effects of sun treatment on $\log \mathrm{LMA}$. This enabled us to disentangle whether $\log \mathrm{LMA}$ effects \aax{} directly or indirectly because both traits respond to growth light intensity. We integrated over missing LMA data using the `mi()` function in **brms**. Because the evidence strong supported including effects of growth and measurement light intensity on the residual variance in $\mathrm{AA}$, we included these in terms in all models. In total, we fit `r n_model2` models, each with a different combination of fixed and random effect structures. We selected the model with the lowest LOOIC value (@tbl-aa_loo2) to generate posterior predictions for hypothesis testing. Parameter estimates from the selected model are shown in @tbl-fit_aa2.

@@

THIS COULD BE USED FOR LMA MODEL? Note that LOOIC was calculated only from pointwise likelihood values of the submodel estimating effects of light intensity, light treatments, and population on \aax{}.

@@

* Table S9: Calculation of alternate scenario where amphi leaves can get same A with less R
* Each A-gs figure with fitted curves, R2, and AA

@@

>>>>

The presence of stomata on both leaf surfaces (amphistomy) can increase photosynthesis in C$_3$ plants by reducing the path length for CO$_2$ diffusion between substomatal cavities and chloroplasts. Amphistomatous leaves are common among herbaceous plants growing in sunny habitats, including many crop species. This distribution of amphistomatous leaves in nature may result from an increased photosynthetic benefit of amphistomy under high light intensity, either because of acclimatory, plastic, or constitutive variation in CO$_2$ supply or demand. We used a recently developed method to quantify amphistomy advantage, the photosynthetic rate of an amphistomatous leaf relative to an otherwise identical hypostomatous leaf, in `r n_acc` diverse populations representing `r n_sp` species of wild tomatoes (*Solanum* sect. *Lycopersicon* and sect. *Lycopersicoides*). Plants grown under high light intensity benefit more in terms of CO$_2$ assimilated for a given stomatal conductance than plants grown under low light, regardless of measurement light intensity. Furthermore, populations native to open habitats tend to benefit more from amphistomy compared to those from more closed habitats. Developmental plasticity and adaptation of leaves to high light intensity likely play a combined role in explaining the adaptive significance and distribution of amphistomy in nature. We estimate that hypostomatous leaves would expend 10-65\% more water to achieve the same photosynthetic rate as a comparable amphistomatous leaf, implying large fitness costs of amphistomy in  ecological contexts where this trait is rare.

@@

The magnitude of \aax{} we document across populations is also noteworthy, as it implies that hypostomatous leaves predominant in mesic to wet forests globally are giving up 'free' carbon that would require little to no additional water loss. The cost of hypostomy can be quantified as the difference in total \gsw{} that an amphistomatous leaf would require to achieve the same photosynthetic rate. This cost can be locally approximated on a log-ratio scale as $\mathrm{AA} / \varepsilon_g$ (Supporting Information), where $\varepsilon_g$ is the elasticity of $A$ to \gsw. For $\mathrm{AA} = 0.05$, the water cost of hypostomy would be 0.5 (64.9\%) at low elasticities ($\varepsilon_g = 0.1$) and 0.1 (10.5\%) at high elasticities ($\varepsilon_g = 0.5$). For plants to expend this much extra water implies a large fitness cost of upper stomata in certain ecological contexts. As this is the first quantitative comparison of \aax{} across species, estimation from a broader range of species and environments will be required to fully understand the ecological causes and consequences of stomatal distributions on leaf surfaces.

@@

To test whether LMA mediated the relationship between growth light intensity and \aax{}, we repeated a similar model comparison procedure but included $\log \mathrm{LMA}$ as an additional fixed effect in the model, as well its interaction with measurement and growth light intensity. The model included a direct effect of growth light intensity on $\log \mathrm{LMA}$ and random phylogeneticlly structured effects of population on $\log \mathrm{LMA}$ and effects of sun treatment on $\log \mathrm{LMA}$. This enabled us to disentangle whether $\log \mathrm{LMA}$ effects \aax{} directly or indirectly because both traits respond to growth light intensity. We integrated over missing LMA data using the `mi()` function in **brms**. Because the evidence strong supported including effects of growth and measurement light intensity on the residual variance in $\mathrm{AA}$, we included these in terms in all models. In total, we fit `r n_model2` models, each with a different combination of fixed and random effect structures. We selected the model with the lowest LOOIC value (@tbl-aa_loo2) to generate posterior predictions for hypothesis testing. Note that LOOIC was calculated only from pointwise likelihood values of the submodel estimating effects of light intensity, light treatments, and population on \aax{}. 
Parameter estimates from the selected model are shown in @tbl-fit_aa2.

@@

### Estimating the water cost of hypostomy

Amphistomy advantage implies that hypostomatous leaves would need to lose more water by having higher \gsw{} to achieve the same photosynthetic rate. We estimated this cost of hypostomy approximately. For simpliticity, we focus on a sufficiently small region of the \agcurve{} curve such that we can approximate the relationship between \gsw{} and $A$ as log-log linear. We also assume that slope is same in both amphistomatous and hypostomatous leaves, which was often approximately the case in our experiment (@fig-curves). With these assumptions, the approximate water cost of hypostomy can be derived from the equations:

$$\log(A_{\mathrm{amphi}}) = \beta_\text{amphi} + \varepsilon_g \log(g_\text{sw}),~\text{and}$$
$$\log(A_{\mathrm{hypo}}) = \beta_\text{hypo} + \varepsilon_g \log(g_\text{sw}).$$
Solving for the difference in $\log{g_\text{sw}}$ at a given $\log{A_{\mathrm{hypo}}}$ is:

$$\log(g_\text{sw,amphi}) - \log(g_\text{sw,hypo}) = \frac{\beta_\text{amphi} - \beta_\text{hypo}}{\varepsilon_g}.$$
Noting that because the curves are parallel, $\mathrm{AA} = \log(A_\text{amphi} /A_\text{hypo}) = \beta_\text{amphi} - \beta_\text{hypo}$, this reveals that the negative log-ratio of the \gsw{} is related to \aax{} divided by the slope:

$$-\log\left(\frac{g_\text{sw,amphi}}{g_\text{sw,hypo}}\right) = \frac{\mathrm{AA}}{\varepsilon_g}.$$
We note that the slope in this case is an elasticity of of $A$ to \gsw{} because both are log-transformed. This quantity will be proportional the water cost of hypostomy assuming that VPD and other environmental variables are held constant and boundary layer conductance are high enough to be negligible.

<!-- estimated the average value of this quantity for each population under each measurement light treatment by regressing \gsw{} against $A$ on a log-log scale, with curve type (amphistomy versus pseudohypostomy) and replicate within population as a random effect using **brms**. We followed the fitting procedures described above for other models. Because of the different assumptions, the estimates of \aax{} in this model will not be identical to that described above, but this approximate method enables a simpler, back-of-the-envelope calculation of the water cost of hypostomy. -->

@@

@@

```{r, curves, echo=FALSE,message=FALSE,warning=FALSE}
#| label: fig-curves
#| fig-cap: "(Figures in separate file) The \\agcurve{} curves, fitted lines, and \\aax{} estimates for every individual plant. The title provides the Tomato Genetics Resource Center accession number, replicate letter, and species names. The subtitle indicates the growth light intensity (sun or shade). Points are raw data, lines are fitted curves, and shaded regions are 95% confidence ribbons for the fitted curves. The Bayesian correlation coefficient $r^2$ is shown to the right of each curve, along with the estimate of \\aax{} $\\pm$ two standard errors for each pair of amphi (orange) and pseudohypo (blue) curves."
ggplot()
```

### 2025-07-27 ms/ms.tex
draft sent for feedback

The presence of stomata on both leaf surfaces (amphistomy) can increase
photosynthesis in C\(_3\) plants by reducing the path length for
CO\(_2\) diffusion between substomatal cavities and chloroplasts.
Amphistomatous leaves are common among herbaceous plants growing in
sunny habitats, including many crop species. This distribution of
amphistomatous leaves in nature may result from an increased
photosynthetic benefit of amphistomy under high light intensity, either
because of acclimatory, plastic, or constitutive variation in CO\(_2\)
supply or demand. We used a recently developed method to quantify
amphistomy advantage, the photosynthetic rate of an amphistomatous leaf
relative to an otherwise identical hypostomatous leaf, in 29 diverse
populations representing 14 species of wild tomatoes (\emph{Solanum}
sect. \emph{Lycopersicon} and sect. \emph{Lycopersicoides}). Plants
grown under high light intensity benefit more in terms of CO\(_2\)
assimilated for a given stomatal conductance than plants grown under low
light, regardless of measurement light intensity. Furthermore,
populations native to open habitats tend to benefit more from amphistomy
compared to those from more closed habitats. Developmental plasticity
and adaptation of leaves to high light intensity likely play a combined
role in explaining the adaptive significance and distribution of
amphistomy in nature.

@@

\subsubsection{MARKER - where is evidence for relationship between PAI
and LMA,
AA?}\label{marker---where-is-evidence-for-relationship-between-pai-and-lma-aa}

The pattern of \aax{} across wild tomatoes strongly supports the
plasticity hypothesis, contradicts the acclimation hypothesis, and
provides modest support for the constitutive hypothesis. Plastic changes
in leaf thickness and/or packing density, summarized by the bulk leaf
mass per area (LMA), may mediate the effect of growth light intensity on
\aax{}. LMA increased in sun grown plants in all populations by an
average of 123\% {[}95\% CI: 42.9 to 256\%{]}, qualitatively similar to
plastic responses in many species (\emph{47}). While LMA is weakly,
albeit significantly (Table~\ref{tbl-fit_aa2}), associated with
individual-level \aax{} (Fig.~\ref{fig-lma_aa}), the effect of growth
light intensity on \aax{} is still predictive based on model comparison
(Table~\ref{tbl-aa_loo1}). In contrast, LMA did not mediate an indirect
effect of native PAI on \aax{} (Table/Figure). Many anatomical traits
underlie LMA (\emph{52}) and future research will be needed identify
which particular traits, such as leaf thickness or mesophyll porosity,
are responsible for mediating \aax{}. The fact that the \aax{} of sun
plants was greater under low measurement light intensity supports a
long-standing hypothesis that resistance to CO\(_2\) diffusion is
greater in the upper than lower portions of the leaf interior
(\emph{53}). At low light intensity, the bulk properties of the leaf are
weighed toward the upper palisade where most light is intercepted.
Hence, amphistomy may, unexpectedly, be particularly beneficial for sun
leaves experiencing intermittent shade or cloud cover. Our study is
limited in testing this because we could not directly measure the
stomatal conductance ratio and intercellular resistance on each surface.
Future experiments measuring \aax{} with a dual sided chamber
(\emph{31}, \emph{33}) can overcome these limitations.

@@

We conclude that developmental plasticity, as opposed to acclimation or
adaptation to open habitats, may explain the long-standing observation
that amphistomatous leaves are more common in sunny habitats, at least
among herbaceous plants including crop relatives. Our results advance
existing explanations by showing high light intensity \emph{per se} does
not increase the benefit of amphistomy, but rather that anatomical and
biochemical change associated with higher light intensity modulate
\aax{}. Developmental plasticity needs to be considered when
interpreting phylogenetic comparative analyses of adaptation, because
plastic differences may be mistaken for genetic differences.

maybe this wording is better: Amphistomy likely plays a significant, but
under appreciated, role in the carbon and water balance of plant groups
where this trait is common, such as herbaceous plants in sunny habitats
and woody species in arid biomes (cites). If our estimates of AA from
the wild tomato clades

note: estimates of AA are less than that described by Parkhurst and Mott
(1990) and also less than that of Xiong and Flexas (2021). What does
this tell us?

A second major conclusion is that the significant photosynthetic gain of
amphistomatous leaves implies that the many hypostomatous leaves that
dominate mesic to wet forests globally are giving up `free' carbon that
would require little to no additional water loss. Under realistic
assumptions, we estimate that hypostomatous leaves could achieve the
same \(A\) with X to X\% less water loss if they were amphistomatous
(Table S9). Ours is the first experimental demonstration that amphistomy
is beneficial for sun leaves because of plastic changes, but future
research needs to estimate the tradeoffs associated with amphistomatous
leaves in order to fully explain their rarity among many woody plants
and shade-tolerant herbs.

@@

To test whether LMA mediated the relationship between growth light
intensity and \aax{}, we repeated a similar model comparison procedure
but included \(\log \mathrm{LMA}\) as an additional fixed effect in the
model, as well its interaction with measurement and growth light
intensity. The model included a direct effect of growth light intensity
on \(\log \mathrm{LMA}\) and random phylogeneticlly structured effects
of population on \(\log \mathrm{LMA}\) and effects of sun treatment on
\(\log \mathrm{LMA}\). This enabled us to disentangle whether
\(\log \mathrm{LMA}\) effects \aax{} directly or indirectly because both
traits respond to growth light intensity. We integrated over missing LMA
data using the \texttt{mi()} function in \textbf{brms}. Because the
evidence strong supported including effects of growth and measurement
light intensity on the residual variance in \(\mathrm{AA}\), we included
these in terms in all models. In total, we fit 75 models, each with a
different combination of fixed and random effect structures. We selected
the model with the lowest LOOIC value (Table~\ref{tbl-aa_loo2}) to
generate posterior predictions for hypothesis testing. Parameter
estimates from the selected model are shown in Table~\ref{tbl-fit_aa2}.

@@

THIS COULD BE USED FOR LMA MODEL? Note that LOOIC was calculated only
from pointwise likelihood values of the submodel estimating effects of
light intensity, light treatments, and population on \aax{}.

@@

\begin{itemize}
\tightlist
\item
  Table S9: Calculation of alternate scenario where amphi leaves can get
  same A with less R
\item
  Each A-gs figure with fitted curves, R2, and AA
\end{itemize}

@@

>>>>

The presence of stomata on both leaf surfaces (amphistomy) can increase
photosynthesis in C\(_3\) plants by reducing the path length for
CO\(_2\) diffusion between substomatal cavities and chloroplasts.
Amphistomatous leaves are common among herbaceous plants growing in
sunny habitats, including many crop species. This distribution of
amphistomatous leaves in nature may result from an increased
photosynthetic benefit of amphistomy under high light intensity, either
because of acclimatory, plastic, or constitutive variation in CO\(_2\)
supply or demand. We used a recently developed method to quantify
amphistomy advantage, the photosynthetic rate of an amphistomatous leaf
relative to an otherwise identical hypostomatous leaf, in 29 diverse
populations representing 14 species of wild tomatoes (\emph{Solanum}
sect. \emph{Lycopersicon} and sect. \emph{Lycopersicoides}). Plants
grown under high light intensity benefit more in terms of CO\(_2\)
assimilated for a given stomatal conductance than plants grown under low
light, regardless of measurement light intensity. Furthermore,
populations native to open habitats tend to benefit more from amphistomy
compared to those from more closed habitats. Developmental plasticity
and adaptation of leaves to high light intensity likely play a combined
role in explaining the adaptive significance and distribution of
amphistomy in nature. We estimate that hypostomatous leaves would expend
10-65\% more water to achieve the same photosynthetic rate as a
comparable amphistomatous leaf, implying large fitness costs of
amphistomy in ecological contexts where this trait is rare.

@@

The pattern of \aax{} across wild tomatoes strongly supports the
plasticity hypothesis, contradicts the acclimation hypothesis, and
provides modest support for the constitutive hypothesis. Plastic changes
in leaf thickness and/or packing density, summarized by the bulk leaf
mass per area (LMA), may mediate the effect of growth light intensity on
\aax{}. LMA increased in sun grown plants in all populations by an
average of 123\% {[}95\% CI: 42.9 to 256\%{]}, qualitatively similar to
plastic responses in many species (\emph{47}). While LMA is weakly,
albeit significantly (Table~\ref{tbl-fit_aa2}), associated with
individual-level \aax{} (Fig.~\ref{fig-lma_aa}), the effect of growth
light intensity on \aax{} is still predictive based on model comparison
(Table~\ref{tbl-aa_loo1}). However, the direct effect of the sun
treatment on \aax{} was weaker when LMA was included in the model
(Table~\ref{tbl-fit_aa1} vs. Table~\ref{tbl-fit_aa2}), suggestive of a
mediating role. Many anatomical traits underlie LMA (\emph{52}) and
future research will be needed identify which particular traits, such as
leaf thickness or mesophyll porosity, are responsible for mediating
\aax{}. The fact that the \aax{} of sun plants was greater under low
measurement light intensity supports a long-standing hypothesis that
resistance to CO\(_2\) diffusion is greater in the upper than lower
portions of the leaf interior (\emph{53}). At low light intensity, the
bulk properties of the leaf are weighed toward the upper palisade where
most light is intercepted. Hence, amphistomy may, unexpectedly, be
particularly beneficial for sun leaves experiencing intermittent shade
or cloud cover. Our study is limited in testing this because we could
not directly measure the stomatal conductance ratio and intercellular
resistance on each surface. Future experiments measuring \aax{} with a
dual sided chamber (\emph{31}, \emph{33}) can overcome these
limitations.

@@

We conclude that both developmental plasticity and adaptation to open
habitats, but not acclimation, may explain the long-standing observation
that amphistomatous leaves are more common in sunny habitats, at least
among herbaceous plants including crop relatives. This result changes
our understanding by showing high light intensity \emph{per se} does not
increase the benefit of amphistomy, but rather arise as a byproduct of
anatomical and biochemical changes caused by plasticity and adaptation
to higher light intensity modulate \aax{}. To gain more precise
understanding of when amphistomy is most beneficial will require further
research on its leaf anataomical and biochemical basis.

The magnitude of \aax{} we document across populations is also
noteworthy, as it implies that hypostomatous leaves predominant in mesic
to wet forests globally are giving up `free' carbon that would require
little to no additional water loss. The cost of hypostomy can be
quantified as the difference in total \gsw{} that an amphistomatous leaf
would require to achieve the same photosynthetic rate. This cost can be
locally approximated on a log-ratio scale as
\(\mathrm{AA} / \varepsilon_g\) (Supporting Information), where
\(\varepsilon_g\) is the elasticity of \(A\) to \gsw. For
\(\mathrm{AA} = 0.05\), the water cost of hypostomy would be 0.5
(64.9\%) at low elasticities (\(\varepsilon_g = 0.1\)) and 0.1 (10.5\%)
at high elasticities (\(\varepsilon_g = 0.5\)). For plants to expend
this much extra water implies a large fitness cost of upper stomata in
certain ecological contexts. As this is the first quantitative
comparison of \aax{} across species, estimation from a broader range of
species and environments will be required to fully understand the
ecological causes and consequences of stomatal distributions on leaf
surfaces.

@@

To test whether LMA mediated the relationship between growth light
intensity and \aax{}, we repeated a similar model comparison procedure
but included \(\log \mathrm{LMA}\) as an additional fixed effect in the
model, as well its interaction with measurement and growth light
intensity. The model included a direct effect of growth light intensity
on \(\log \mathrm{LMA}\) and random phylogeneticlly structured effects
of population on \(\log \mathrm{LMA}\) and effects of sun treatment on
\(\log \mathrm{LMA}\). This enabled us to disentangle whether
\(\log \mathrm{LMA}\) effects \aax{} directly or indirectly because both
traits respond to growth light intensity. We integrated over missing LMA
data using the \texttt{mi()} function in \textbf{brms}. Because the
evidence strong supported including effects of growth and measurement
light intensity on the residual variance in \(\mathrm{AA}\), we included
these in terms in all models. In total, we fit 75 models, each with a
different combination of fixed and random effect structures. We selected
the model with the lowest LOOIC value (Table~\ref{tbl-aa_loo2}) to
generate posterior predictions for hypothesis testing. Note that LOOIC
was calculated only from pointwise likelihood values of the submodel
estimating effects of light intensity, light treatments, and population
on \aax{}. Parameter estimates from the selected model are shown in
Table~\ref{tbl-fit_aa2}.

@@

\subsubsection{Estimating the water cost of
hypostomy}\label{estimating-the-water-cost-of-hypostomy}

Amphistomy advantage implies that hypostomatous leaves would need to
lose more water by having higher \gsw{} to achieve the same
photosynthetic rate. We estimated this cost of hypostomy approximately.
For simpliticity, we focus on a sufficiently small region of the
\agcurve{} curve such that we can approximate the relationship between
\gsw{} and \(A\) as log-log linear. We also assume that slope is same in
both amphistomatous and hypostomatous leaves, which was often
approximately the case in our experiment (Figure~\ref{fig-curves}). With
these assumptions, the approximate water cost of hypostomy can be
derived from the equations:

\[\log(A_{\mathrm{amphi}}) = \beta_\text{amphi} + \varepsilon_g \log(g_\text{sw}),~\text{and}\]
\[\log(A_{\mathrm{hypo}}) = \beta_\text{hypo} + \varepsilon_g \log(g_\text{sw}).\]
Solving for the difference in \(\log{g_\text{sw}}\) at a given
\(\log{A_{\mathrm{hypo}}}\) is:

\[\log(g_\text{sw,amphi}) - \log(g_\text{sw,hypo}) = \frac{\beta_\text{amphi} - \beta_\text{hypo}}{\varepsilon_g}.\]
Noting that because the curves are parallel,
\(\mathrm{AA} = \log(A_\text{amphi} /A_\text{hypo}) = \beta_\text{amphi} - \beta_\text{hypo}\),
this reveals that the negative log-ratio of the \gsw{} is related to
\aax{} divided by the slope:

\[-\log\left(\frac{g_\text{sw,amphi}}{g_\text{sw,hypo}}\right) = \frac{\mathrm{AA}}{\varepsilon_g}.\]
We note that the slope in this case is an elasticity of of \(A\) to
\gsw{} because both are log-transformed. This quantity will be
proportional the water cost of hypostomy assuming that VPD and other
environmental variables are held constant and boundary layer conductance
are high enough to be negligible.

@@

@@

\pandocbounded{\includegraphics[keepaspectratio]{ms_files/figure-pdf/fig-curves-1.pdf}}

\caption{\label{fig-curves}(Figures in separate file) The \agcurve{}
curves, fitted lines, and \aax{} estimates for every individual plant.
The title provides the Tomato Genetics Resource Center accession number,
replicate letter, and species names. The subtitle indicates the growth
light intensity (sun or shade). Points are raw data, lines are fitted
curves, and shaded regions are 95\% confidence ribbons for the fitted
curves. The Bayesian correlation coefficient \(r^2\) is shown to the
right of each curve, along with the estimate of \aax{} \(\pm\) two
standard errors for each pair of amphi (orange) and pseudohypo (blue)
curves.}
